Semantics for the $ \lambda $-calculus

### 2024-05-16 report/chapters/category-theory.tex
Write about the karoubi envelope of a Lambda-algebra

@@

>>>>

Now, if we have an object in a slice category, we can again look at the slice of the slice category over that object. However, this gives us nothing new:
\begin{lemma}
  Let $ (d, f) $ be an object in the slice category $ (C \downarrow c) $. The slice category $ ((C \downarrow c) \downarrow (d, f)) $ is equivalent to $ (C \downarrow d) $.
\end{lemma}
\begin{proof}
  An object of $ ((e, g), \alpha) : ((C \downarrow c) \downarrow (d, f)) $ is an object $ (e, g) : (C \downarrow c) $, together with a morphism $ \alpha: (C \downarrow c)((e, g), (d, f)) $. That is, a morphism $ \alpha: C(e, d) $ such that the obvious triangle commutes (shown in the diagram below on the far left).

  Then a morphism between $ ((e, g), \alpha) $ and $ ((e^\prime, g^\prime), \alpha^\prime) $ is a morphism $ \beta $ between $ (e, g) $ and $ (e^\prime, g^\prime) $ that commutes with $ \alpha $ and $ \alpha^\prime $. Note that a morphism between $ (e, g) $ and $ (e^\prime, g^\prime) $ is a morphism between $ e $ and $ e^\prime $ that commutes with $ g $ and $ g^\prime $.
  \begin{center}
    \begin{tikzcd}
      e \arrow[rdd, "g"'] \arrow[rd, "\alpha"] \arrow[rr, "\beta"] && e^\prime \arrow[ldd, "g^\prime"] \arrow[ld, "\alpha^\prime"']\\
      & d \arrow[d, "f" description]\\
      & c
    \end{tikzcd}
    $ \Leftrightarrow $
    \begin{tikzcd}
      e \arrow[rd, "\alpha"] \arrow[rr, "\beta"] && e^\prime \arrow[ld, "\alpha^\prime"']\\
      & d
    \end{tikzcd}
  \end{center}

  Now, note that $ g $ and $ g^\prime $ are completely determined by $ g = \alpha \cdot f $ and $ g^\prime = \alpha^\prime \cdot f $, so we can leave them out. Also, if $ \beta $ commutes with $ \alpha $ and $ \alpha^\prime $, it automatically also commutes with $ g $ and $ g^\prime $. Therefore, as shown above, we have a correspondence between objects and morphisms
  \[ \beta: ((e, g), \alpha) \to ((e^\prime, g^\prime), \alpha^\prime) \Leftrightarrow \beta: (e, \alpha) \to (e^\prime, \alpha^\prime). \]
\end{proof}

Also, it turns out that we can derive some structure of the slice categories from the original category:
\begin{lemma}
  For $ (d, f), (d^\prime, f^\prime) : (C \downarrow c) $, their product in the slice category is given by their pullback / fibered product $ d \times_c d^\prime $, together with the induced morphism $ g: d \times_c d^\prime \dasharrow c $.
\end{lemma}
\begin{proof}
  Consider the diagram below. The fibered product gives `projections' $ h $ and $ h^\prime $. Also, if we have some $ (e, \alpha) : (C \downarrow c) $, together with morphisms $ \beta : (e, \alpha) \to (d, f) $ and $ \beta^\prime : (e, \alpha) \to (d^\prime, f^\prime) $. Then $ \beta $ and $ \beta^\prime $ commute with $ f $ and $ f^\prime $, so by the universal property of the fibred product, there exists a unique morphism $ \gamma : e \to d \times_c d^\prime $ that makes the triangles with $ \beta $ and $ h $, and with $ \beta^\prime $ and $ h^\prime $ commute. Then $ \gamma $ commutes with $ \alpha $ and $ g $ as well, so it is a morphism in $ (C \downarrow c) $. This shows that $ (d \times_c d^\prime, g) $ has the universal property of the product in $ (C \downarrow c) $.
  \begin{center}
    \begin{tikzcd}
      e \arrow[rd, "\beta"] \arrow[rr, "\beta^\prime", bend left] \arrow[rrd, "\alpha"', bend right=49, shift right] \arrow[r, dashed, "\gamma"] & d \times_c d^\prime \arrow[r, "h^\prime"'] \arrow[d, "h"] \arrow[rd, dashed, "g"] & d^\prime \arrow[d, "f^\prime"] \\
      & d \arrow[r, "f"] & c
    \end{tikzcd}
  \end{center}
\end{proof}

For a category $ C $ with products, Hyland introduces the notation $ \Delta_X Y $ for the element $ (X \times Y, p_1) : (C \downarrow X) $, and we will follow his example in this.

In fact, $ \Delta_X : C \to C / X $ is a functor, with $ \Delta_X(f) = \id X \times f $ for $ f: C(Y, Y^\prime) $. This functor preserves the terminal object, products and pullbacks:
\begin{lemma}\label{lem:delta-limits}
  $ \Delta_X $ preserves all limits.
\end{lemma}
\begin{proof}
  Take a diagram $ ((A_i)_i, (f_j)_j) $ in $ C $. Suppose that this has a limit $ (L, (g_i)_i) : C $. Now, consider an object $ (B, q) : (C \downarrow X) $, together with morphisms $ h_i : (C \downarrow X)((B, q), \Delta_X A_i) $, that commute with the $ \Delta_X f_j $:
  \begin{center}
    \begin{tikzcd}
      & B \arrow[ldd, "h_{m_j}"'] \arrow[rdd, "h_{n_j}"]\\
      & X \times L \arrow[ld, "\Delta_X g_{m_j}"] \arrow[rd, "\Delta_X g_{n_j}"']\\
      X \times A_{m_j} \arrow[rr, "\Delta_X f_j"'] & & X \times A_{n_j}
    \end{tikzcd}
  \end{center}
  Then the morphisms in $ (C \downarrow X)((B, q), \Delta_X L) $, commuting with the $ \Delta_X g_j $ and $ h_j $ are the morphisms in $ C(B, X \times L) \cong C(B, X) \times C(B, L) $ that commute with the projections to $ X $ and the $ \id X \times g_j $ and $ h_j $. Since the morphisms in $ C(B, X) $ commuting with $ q $ and $ \id X $ are exactly $ q $, we can forget about this component, and the morphisms we are looking for correspond to the morphisms in $ C(B, L) $ that commute with the $ g_j $ and $ h_j \cdot p_1 $. Since $ (L, (g_j)_j) $ is a limit, this is a unique morphism.
\end{proof}

\begin{lemma}\label{lem:delta-exponentials}
  $ \Delta_X $ preserves exponential objects:
\end{lemma}
\begin{proof}
  See \cite{borceux-3}, Lemma 5.8.2. This lemma shows this via the following equivalences:
  \begin{align*}
    (C \downarrow X)((A, f), \Delta_X Y^Z) &\cong C(A, Y^Z)\\
    &\cong C(A \times Z, Y)\\
    &\cong (C \downarrow X)((A \times X \times Z, \langle f, p_1 \rangle), \Delta_X Y)\\
    &\cong (C \downarrow X)((A, f) \times \Delta_X Z, \Delta_X Y)\\
  \end{align*}
\end{proof}

@@

\begin{remark}\label{rem:pullback-of-projection}
  Note that for $ X, Y, Z : C $ and $ f: C(Y, X) $, the following diagram shows that $ f^* \Delta_X Z \cong \Delta_Y Z $:
  \begin{center}
    \begin{tikzcd}
      Y \times Z \arrow[d, "p_1"] \arrow[r, "f \times \id Z"] & X \times Z \arrow[d, "p_1"] \arrow[r, "p_2"] & Z \arrow[d, "!"]\\
      Y \arrow[r, "f"] & X \arrow [r, "!"] & T
    \end{tikzcd}
  \end{center}
\end{remark}

\begin{lemma}\label{lem:constant-dependent-product}
  For $ W, X, Z : C $ and $ p_1 : X \times Z \to X $,
  \[ \prod_{p_1} \Delta_{X \times Z} W \cong \Delta_X W^Z \]
\end{lemma}
\begin{proof}
  First of all, note that $ (C \downarrow X \times Z) \cong ((C \downarrow X) \downarrow \Delta_X Z) $. Also note that the composite morphism $ (X \times Z) \times W \xrightarrow{p_1} X \times Z \xrightarrow{p_1} X $ is the element $ \Delta_X (W \times Z) : (C \downarrow X) $.

  By Proposition 1.34 in \cite{freyd}, $ \prod_{p_1} \Delta_{X \times Z} W $ is given as the following pullback:
  \begin{center}
    \begin{tikzcd}
      \prod_{p_1} \Delta_{X \times Z} W \arrow[r] \arrow[d] & (\Delta_X Z \times W)^{\Delta_X Z} \arrow[d, "(\Delta_X p_1)^{\Delta_X Z}"]\\
      X \arrow[r] & (\Delta_X Z)^{\Delta_X Z}
    \end{tikzcd}
  \end{center}

  By Lemma \ref{lem:delta-exponentials}, $ \Delta_X $ preserves exponential objects, so the morphism on the right is $ \Delta_X p_1^Z : (C \downarrow X)(\Delta_X (Z \times W)^Z, \Delta_X Z^Z) $. However, we have an isomorphism $ (Z \times W)^Z \cong Z^Z \times Z^W $, and then the morphism on the right becomes
  \[ \Delta_X p_1 : (C \downarrow X)(\Delta_X(Z^Z \times Z^W), \Delta_X Z^Z) \]
  We also have an isomorphism $ X \cong \Delta_X T $. Then by Lemma \ref{lem:delta-limits} and Remark \ref{rem:pullback-of-projection}, the pullback of this diagram is $ \Delta_X W^Z $:
  \begin{center}
    \begin{tikzcd}
      \Delta_X W^Z \arrow[r] \arrow[d] & \Delta_X (Z^Z \times W^Z) \arrow[d, "\Delta_X p_1"]\\
      \Delta_X T \arrow[r] & \Delta_X Z^Z
    \end{tikzcd}
  \end{center}
\end{proof}

### 2024-05-16 report/chapters/paper.tex
Write about the karoubi envelope of a Lambda-algebra

\begin{lemma}
  The precomposition functors $ i_* $ and $ j_* $ are adjoint equivalences.
\end{lemma}
\begin{proof}
  We will show that $ j_* $ is an adjoint equivalence. From the `two out of three' property, it follows then that $ i_* $ is also an adjoint equivalence.

@@

\begin{lemma}
  The square in the diagram above 2-commutes.
\end{lemma}
\begin{proof}
  Note:
  \[ (Y \bullet j_*)(A)(n) = (j \bullet (Y(A)))(n) = Y(A)(j(n)) = \R(U^n, A) \]
  \[ (\iota \bullet \sim)(A)(n) = \{ a : L_n \mid A a = a \} \]

  By Scott's representation theorem, we have $ \mathbb R(U^n, U) \cong L_n, $ via
  \[ f \mapsto \iota_{0, n}(f) ((((c, x_1), x_2), \dots), x_n). \]
  Restricting this equivalence to the morphisms to $ A $ yields
  \[ \R(U^n, A) \cong \{ a : L_n \mid A a = a \}. \]
  Of course, this is natural in $ A $, since for $ g: \R(A, B) $, postcomposing elements of $ \R(U^n, A) $ by $ g $ is equivalent to applying $ g $ to the elements of $ \{ a : L_n \mid A a = a \} $.

  \TODO: Check naturality in $ n $.
\end{proof}

@@

For a category $ C $ with products, Hyland introduces the notation $ \Delta_X Y $ for the element $ (X \times Y, p_1) : (C \downarrow X) $, and we will follow his example in this.

@@

\begin{remark}\label{rem:pullback-of-projection}
  Note that for $ X, Y, Z : C $ and $ f: C(Y, X) $, the following diagram shows that $ f^* \Delta_X Z \cong \Delta_Y Z $:
  \begin{center}
    \begin{tikzcd}
      Y \times Z \arrow[d, "p_1"] \arrow[r, "f \times \id Z"] & X \times Z \arrow[d, "p_1"] \arrow[r, "p_2"] & Z \arrow[d, "!"]\\
      Y \arrow[r, "f"] & X \arrow [r, "!"] & T
    \end{tikzcd}
  \end{center}
\end{remark}

In fact, $ \Delta_X : C \to C / X $ is a functor, with $ \Delta_X(f) = \id X \times f $ for $ f: C(Y, Y^\prime) $. This functor preserves the terminal object, products and pullbacks:
\begin{lemma}\label{lem:delta-limits}
  $ \Delta_X $ preserves all limits.
\end{lemma}
\begin{proof}
  Take a diagram $ ((A_i)_i, (f_j)_j) $ in $ C $. Suppose that this has a limit $ (L, (g_i)_i) : C $. Now, consider an object $ (B, q) : (C \downarrow X) $, together with morphisms $ h_i : (C \downarrow X)((B, q), \Delta_X A_i) $, that commute with the $ \Delta_X f_j $:
  \begin{center}
    \begin{tikzcd}
      & B \arrow[ldd, "h_{m_j}"'] \arrow[rdd, "h_{n_j}"]\\
      & X \times L \arrow[ld, "\Delta_X g_{m_j}"] \arrow[rd, "\Delta_X g_{n_j}"']\\
      X \times A_{m_j} \arrow[rr, "\Delta_X f_j"'] & & X \times A_{n_j}
    \end{tikzcd}
  \end{center}
  Then the morphisms in $ (C \downarrow X)((B, q), \Delta_X L) $, commuting with the $ \Delta_X g_j $ and $ h_j $ are the morphisms in $ C(B, X \times L) \cong C(B, X) \times C(B, L) $ that commute with the projections to $ X $ and the $ \id X \times g_j $ and $ h_j $. Since the morphisms in $ C(B, X) $ commuting with $ q $ and $ \id X $ are exactly $ q $, we can forget about this component, and the morphisms we are looking for correspond to the morphisms in $ C(B, L) $ that commute with the $ g_j $ and $ h_j \cdot p_1 $. Since $ (L, (g_j)_j) $ is a limit, this is a unique morphism.
\end{proof}

\begin{lemma}\label{lem:delta-exponentials}
  $ \Delta_X $ preserves exponential objects:
\end{lemma}
\begin{proof}
  See \cite{borceux-3}, Lemma 5.8.2. This lemma shows this via the following equivalences:
  \begin{align*}
    (C \downarrow X)((A, f), \Delta_X Y^Z) &\cong C(A, Y^Z)\\
    &\cong C(A \times Z, Y)\\
    &\cong (C \downarrow X)((A \times X \times Z, \langle f, p_1 \rangle), \Delta_X Y)\\
    &\cong (C \downarrow X)((A, f) \times \Delta_X Z, \Delta_X Y)\\
  \end{align*}
\end{proof}

\begin{lemma}\label{lem:constant-dependent-product}
  For $ W, X, Z : C $ and $ p_1 : X \times Z \to X $,
  \[ \prod_{p_1} \Delta_{X \times Z} W = \Delta_X W^Z \]
\end{lemma}
\begin{proof}
  First of all, note that $ (C \downarrow X \times Z) \cong ((C \downarrow X) \downarrow \Delta_X Z) $. Also note that the composite morphism $ (X \times Z) \times W \xrightarrow{p_1} X \times Z \xrightarrow{p_1} X $ is the element $ \Delta_X (W \times Z) : (C \downarrow X) $.

  By Proposition 1.34 in \cite{freyd}, $ \prod_{p_1} \Delta_{X \times Z} W $ is given as the following pullback:
  \begin{center}
    \begin{tikzcd}
      \prod_{p_1} \Delta_{X \times Z} W \arrow[r] \arrow[d] & (\Delta_X Z \times W)^{\Delta_X Z} \arrow[d, "(\Delta_X p_1)^{\Delta_X Z}"]\\
      X \arrow[r] & (\Delta_X Z)^{\Delta_X Z}
    \end{tikzcd}
  \end{center}

  By Lemma \ref{lem:delta-exponentials}, $ \Delta_X $ preserves exponential objects, so the morphism on the right is $ \Delta_X p_1^Z : (C \downarrow X)(\Delta_X (Z \times W)^Z, \Delta_X Z^Z) $. However, we have an isomorphism $ (Z \times W)^Z \xrightarrow \sim Z^Z \times Z^W $, and then the morphism on the right becomes
  \[ \Delta_X p_1 : (C \downarrow X)(\Delta_X(Z^Z \times Z^W), \Delta_X Z^Z) \]
  We also have an isomorphism $ X \cong \Delta_X T $. Then by Lemma \ref{lem:delta-limits} and Remark \ref{rem:pullback-of-projection}, the pullback of this diagram is $ \Delta_X W^Z $:
  \begin{center}
    \begin{tikzcd}
      \Delta_X W^Z \arrow[r] \arrow[d] & \Delta_X (Z^Z \times W^Z) \arrow[d, "p_1"]\\
      \Delta_X T \arrow[r] & \Delta_X Z^Z
    \end{tikzcd}
  \end{center}
\end{proof}

@@

The $ \Lambda $-algebra structure gives the terms of $ A $ quite a lot of behaviour. For example, we can define `function application' as
\[ a b = (x_1 x_2) \bullet (a, b) \]
and composition as
\[ a \circ b = (x_1 \circ x_2) \bullet (a, b) \]
for $ a, b : A $, with $ x_1 \circ x_2 = \lambda x_3, x_1 (x_2 x_3) : \Lambda_2 $.

@@

The fundamental theorem states that there is an adjoint equivalence
\[ \LamTh \cong \Alg \Lambda. \]
We will prove this by showing that there is a weak equivalence.

@@

@@

\begin{definition}[The monoid of a $ \Lambda $-algebra]
  Now we make $ A_1 $ into a monoid under composition $ \circ $ with unit $ \mathbf 1_1 $. The fact that this is a monoid follows from the remarks in the last section.
\end{definition}

@@

\begin{definition}
  Now, composition $ \circ $ gives a right $ A_1 $-action on the $ A_n $, so we have $ A_n : \RAct{A_1} $.
\end{definition}

\begin{lemma}
  We have $ U_{A_1}^{U_{A_1}} \cong A_2 $ in $ \RAct{A_1} $.
\end{lemma}
\begin{proof}
  Recall that $ U_{A_1}^{U_{A_1}} $ consists of the set of $ A_1 $-equivariant morphisms $ U_{A_1} \times U_{A_1} \to U_{A_1} $.

  We have a bijection $ \varphi: A_2 \xrightarrow{\sim} U_{A_1}^{U_{A_1}} $, given by
  \[ \varphi(a)(b, b^\prime) = (\lambda x_4, x_1 (x_2 x_4) (x_3 x_4)) \bullet (a, b, b^\prime), \]
  \TODO: relate to product in category of retracts.

  with an inverse given by
  \[ \varphi^{-1}(f) = \lambda x_1 x_2, f(p_1, p_2)(\lambda x_3, x_3 x_1 x_2) \]
  \TODO: Make more explicitly that we use $ \bullet $ here.
  for $ p_i = \lambda x_1, x_1 (\lambda x_2 x_3, x_{i + 1}) $. Note that for terms $ c_1, c_2 $, we have $ p_i (\lambda x_1, x_1 c_1 c_2) = c_i $.

  This is an inverse, because given $ f: U_{A_1}^{U_{A_1}} $ and $ (a_1, a_2): U_{A_1} \times U_{A_1} $, we have
  \[ \varphi(\psi(f))(a_1, a_2) = f(p_1, p_2) \circ q = f(p_1 \circ q, p_2 \circ q) = f(a_1, a_2) \]
  for $ q = \lambda x_1, (\lambda x_2, x_2 (a_1 x_1) (a_2 x_1)) $. In the last step of this proof, we use, among other things, the fact that the $ a_i : A_1 $ and therefore $ \lambda x_1, a_i x_1 = a_i $.

  Some straightforward rewriting shows that for $ a: A_2 $, we have $ \psi(\varphi(a)) = a $. In the last step of this proof, we use the fact that $ a : A_2 $ and therefore $ \lambda x_1 x_2, a x_1 x_2 = a $.

  Therefore, $ \varphi $ is a bijection and, as it turns out, an isomorphism.
\end{proof}

\begin{definition}[Construction of the $ \lambda $-theory]
  Since $ \RAct{A_1} $ has products, the algebraic theory $ E(U_{A_1}) $ exists.

  Recall that $ A_2 \subseteq A_1 $ and that $ a \mapsto \mathbf 1_2 \circ a $ gives a function from $ A_1 $ to $ A_2 $, which is, by definition, the identity on $ A_2 $. This gives $ E(U_{A_1}) $ a $ \lambda $-theory structure with $ \beta $-equality.
\end{definition}

>>>>

\begin{lemma}\label{lem:equivalent-presheaf-cats}
  The precomposition functors $ i_* $ and $ j_* $ are adjoint equivalences.
\end{lemma}
\begin{proof}
  We will show that $ j_* $ is an adjoint equivalence. From the `two out of three' property, it follows then that $ i_* $ is also an adjoint equivalence.

@@

\begin{lemma}
  The square in the diagram above 2-commutes.
\end{lemma}
\begin{proof}
  Note:
  \[ (Y \bullet j_*)(A)(n) = (j \bullet (Y(A)))(n) = Y(A)(j(n)) = \R(U^n, A) \]
  \[ (\iota \bullet \sim)(A)(n) = \{ a : L_n \mid \iota_{0, n}(A) a = a \} \]

  By Scott's representation theorem, we have $ \varphi_n: \R(U^n, U) \xrightarrow \sim L_n, $ given by
  \[ \varphi_n(f) = \iota_{0, n}(f) ((((c, x_1), x_2), \dots), x_n). \]
  Restricting this equivalence to the morphisms to $ A $ yields
  \[ \R(U^n, A) \cong \{ a : L_n \mid A a = a \}. \]
  Of course, this is natural in $ A $, since for $ g: \R(A, B) $, postcomposing elements of $ \R(U^n, A) $ by $ g $ is equivalent to applying $ g $ to the elements of $ \{ a : L_n \mid A a = a \} $.

  Lastly, for $ f : L_m^n $, the presheaf actions
  \[ (Y \bullet j_*)(A)(f) : \R(U^n, A) \to \R(U^m, A) \]
  and
  \[ (\iota \bullet \sim)(A)(f) : \{ a : L_n \mid \iota_{0, n}(A) a = a \} \to \{ a : L_m \mid \iota_{0, n}(A) a = a \} \]
  are in fact given by the substitution operations $ \bullet $ in $ E(U) $ and $ L $. Since Scott's representation theorem shows that $ E(U) $ is isomorphic to $ L $ as a $ \lambda $-theory, these substitutions are compatible with the $ \varphi_n $, which shows naturality in $ n $.
\end{proof}

@@

@@

@@

The $ \Lambda $-algebra structure gives the terms of $ A $ quite a lot of behaviour. For example, we can define `function application' as
\[ a b = (x_1 x_2) \bullet (a, b) \]
and composition as
\[ a \circ b = (x_1 \circ x_2) \bullet (a, b) \]
and the same for the other constructions at the start of Section \ref{sec:retracts-category}.

@@

The fundamental theorem states that there is an adjoint equivalence
\[ \LamTh \cong \Alg \Lambda. \]
We will prove this by showing that there is a weak equivalence. Using the fact that $ \LamTh $ is

@@

Here, we are mostly interested in the case $ n = 0 $.

@@

\begin{definition}[The monoid of a $ \Lambda $-algebra]
  We make $ A_1 $, the `functional elements' of $ A $, into a monoid under composition $ \circ $ with unit $ \mathbf 1_1 $. The fact that this is a monoid follows from the remarks in the last section.
\end{definition}

@@

\begin{remark}
  Note that, like in the last chapter, we can take the Karoubi envelope $ \R_A $ of $ C_{A_1} $, which yields the following diagram:
  \begin{center}
    \begin{tikzcd}
      & C_{A_1} \arrow[d, hookrightarrow, "Y"] \arrow[r, hookrightarrow, "\iota_{C_{A_1}}"] & \R_A \arrow[d, hookrightarrow, "Y"]\\
      \RAct{A_1} & P C_{A_1} \arrow[l, "\sim"] & P \R_A \arrow[l, "\sim"]
    \end{tikzcd}
  \end{center}
  Explicitly, this gives the embedding $ \R_A(U, -): \R_A \hookrightarrow \RAct{A_1} $ given by
  \[ X \mapsto \R_A(U, X) = \{ x : A \mid X \circ x = x \}. \]
  Actually, we can repeat almost the entirety of Section \ref{sec:retracts-category} for $ \R_A $. In particular, $ \R_A $ has products
  \[ A \times B = \langle x_1 \circ \pi_1, x_2 \circ \pi_2 \rangle \bullet (A, B) \]
  and exponential objects
  \[ B^A = (\lambda x_3, x_1 \circ x_3 \circ x_2) \bullet (B, A) \]
  with isomorphism
  \[ \psi: \R_A(C \times A, B) \xrightarrow \sim \R_A(C, B^A) \]
  given by
  \[ \psi(f) = (\lambda x_2, x_1 (\pi_1 x_2) (\pi_2 x_2)) \bullet f \quad \text{and} \quad \psi^{-1}(f) = (\lambda x_2 x_3, x_1 (x_2, x_3)) \bullet f. \]
\end{remark}

\begin{remark}
  Also, note that if $ A = L_0 $ for some $ \lambda $-theory $ L $, then the monoid $ A_1 $ is equivalent to the monoid $ L_1 $, and $ \R_A $ is equivalent to our familiar category of retracts $ \R $. From Lemma \ref{lem:equivalent-presheaf-cats}, it follows that $ \RAct{A_1} $ is equivalent to $ \Pshf L $. This might motivate some of the following constructions.
\end{remark}

\begin{definition}
  Now, composition $ \circ $ gives a right $ A_1 $-action on the $ A_n $, so we have $ A_n : \RAct{A_1} $ (note that this is the image of $ \mathbf 1_n : \R_A $ under the embeddding into $ \RAct{A_1} $).
\end{definition}

\begin{lemma}
  $ A_2 $ is the exponential object $ A_1^{A_1} $ in $ \RAct{A_1} $.
\end{lemma}
\begin{proof}
  Note that the embedding $ \R_A \hookrightarrow \RAct{A_1} $ is the composition of a Yoneda embedding and two adjoint equivalences. These all preserve exponential objects (\TODO: Citation needed). Since $ A_1 $ is the image of $ \mathbf 1_1 : \R_A $, the exponential $ A_1^{A_1} $ is $ A_2 $, the image of $ \mathbf 1_1^{\mathbf 1_1} = \lambda x_1 x_2, x_1 x_2 $.
\end{proof}

\begin{remark}
  Hyland shows this by explicitly constructing an isomorphism $ \psi $ between $ A_2 $ and the `familiar' exponential object in $ \RAct{A_1} $ (the set of $ A_1 $-equivariant morphisms $ U_{A_1} \times U_{A_1} \to U_{A_1} $), given by
  \[ \varphi(f) = (b, b^\prime) \mapsto (\lambda x_4, x_1 (x_2 x_4) (x_3 x_4)) \bullet (f, b, b^\prime), \]
  treating $ f : A_2 $ as a function in two elements, and simultaneously composing it with the functions $ b, b^\prime : A_1 $ in one argument. It has an inverse
  \[ \varphi^{-1}(g) = (\lambda x_2 x_3, x_1 (x_2, x_3)) \bullet (g(\pi_1, \pi_2)). \]
  Note that the first pair $ (x_2, x_3) $ is a $ \lambda $-term, whereas the second pair $ (\pi_1, \pi_2) $ is a pair in $ U_{A_1} \times U_{A_1} $.
\end{remark}

\begin{definition}[Construction of the $ \lambda $-theory]
  Now, note that, like before, $ \mathbf 1_1 $ is a retract of $ \mathbf 1_2 $ in $ \R_A $. The embedding into $ \RAct{A_1} $ turns this into the embedding as a subset $ A_2 \subseteq A_1 $ and a retraction $ r: A_1 \to A_2 $, given by precomposition with $ \mathbf 1_2 $.

  Therefore, $ U_{A_1} $ is a reflexive object in $ \RAct{A_1} $, and we get an $ \lambda $-theory $ E(U_{A_1}) $.
\end{definition}

### 2024-05-16 report/chapters/previous-work.tex
Write about the karoubi envelope of a Lambda-algebra

\begin{tabular}{|c|c|c|}\hline
  $ K $ & $ \lambda xy.x $ & constant \\\hline
  $ S $ & $ \lambda xyz.xz(yz) $ & \\\hline
  $ I $ & $ \lambda x.x = SKK $ & identity \\\hline
  $ Y $ & $ \lambda f.(\lambda x.f (xx))(\lambda x.f (xx)) $ & ﬁxpoint \\\hline
  $ \bot $ & $ (\lambda x.xx)(\lambda x.xx) = (SII)(SII) = YI $ & bottom \\\hline
  $ P $ & $ \lambda f gx.g(f x) = S(K(S(S(KS)K)))K $ & composition \\\hline
  $ Q $ & $ \lambda f gax.ga(f ax) $ & ﬁbred composition \\\hline
  $ \langle \rangle $ & $ \lambda xyz.zxy $ & pairing \\\hline
  $ 0 $ & $ \lambda xy.x = K $ & left projection \\\hline
  $ 1 $ & $ \lambda xy.y = KI $ & right projection \\\hline
\end{tabular}

@@

\section{The category of retracts}

@@

\begin{remark}
  Hyland instead defines $ \R $ as
  \[ \R_0 = \{ A : L_1 \mid A \bullet A = A \} \quad \text{and} \R(A, B) = \{ f : L_1 \mid B \bullet f \bullet A = f \}, \]
  writing $ s \bullet t $ instead of $ s \bullet (t) $ for $ s, t : L_1 $. Note that the following diagram commutes:
  \begin{center}
    \begin{tikzcd}
      L_1 \times L_1 \arrow[r, "\bullet"] \arrow[d, "{(s, t) \mapsto ((\lambda x_1, s), (\lambda x_1, t))}"'] & L_1 \arrow[d, "{s \mapsto (\lambda x_1, s)}"]\\
      L_0 \times L_0 \arrow[r, "\circ"] & L_0
    \end{tikzcd}
  \end{center}
  Now, since for $ A : \R $, $ A = \lambda x_1, \iota_{0, 1}(A) (\iota_{0, 1}(A) x_1) $, we have $ \lambda x_1, \iota_{0, 1}(A) x_1 = A $. This shows that $ s \mapsto \lambda x_1, s $ and $ t \mapsto \iota_{0, 1}(t) x_1 $ (both on objects and morphisms) constitute an equivalence between Hyland's category of retracts and Scott's category of retracts.
\end{remark}

@@

We might wonder whether $ \R $ also has an initial object $ O $. However, for any $ c: L_0 $, we would have a constant morphism
\[ I_c = \lambda x_1, \iota_{0, 1}(c) : \R(A, U), \]
so if $ L $ is nontrivial, $ \R $ has no initial object.

@@

$ \R $ also has exponential objects
\[ C^B = \lambda x_1, C \circ x_1 \circ B \]
with evaluation morphism $ \epsilon_{BC}: C^B \times B \to C $ given by
\[ \epsilon_{BC} = \lambda x_1, C(\pi_1 x_1 (B (\pi_2 x_1))), \]
which is universal because we can lift a morphism $ f: \R(A \times B, C) $ to a morphism $ \psi(f): \R(A, C^B) $ given by
\[ \psi(f) = \lambda x_1 x_2, f (x_1, x_2). \]
Note that for $ g: \R(A, C^B) $, the inverse $ \psi^{-1}(g) $ is given by
\[ \epsilon \circ \langle g \circ \pi_1, B \circ \pi_2 \rangle = \lambda x_1, g (\pi_1 x_1) (\pi_2 x_1): \R(A \times B, C). \]
Also note that
\[ \lambda x_1, C \circ x_1 \circ B = C^B = \id{C^B} = \psi(\epsilon_{BC}). \]
The first representation shows that $ \varphi(C^B)_0 = \R(B, C) $, as we would expect.

>>>>

% \begin{tabular}{|c|c|c|}\hline
%   $ K $ & $ \lambda xy.x $ & constant \\\hline
%   $ S $ & $ \lambda xyz.xz(yz) $ & \\\hline
%   $ I $ & $ \lambda x.x = SKK $ & identity \\\hline
%   $ Y $ & $ \lambda f.(\lambda x.f (xx))(\lambda x.f (xx)) $ & ﬁxpoint \\\hline
%   $ \bot $ & $ (\lambda x.xx)(\lambda x.xx) = (SII)(SII) = YI $ & bottom \\\hline
%   $ P $ & $ \lambda f gx.g(f x) = S(K(S(S(KS)K)))K $ & composition \\\hline
%   $ Q $ & $ \lambda f gax.ga(f ax) $ & ﬁbred composition \\\hline
%   $ \langle \rangle $ & $ \lambda xyz.zxy $ & pairing \\\hline
%   $ 0 $ & $ \lambda xy.x = K $ & left projection \\\hline
%   $ 1 $ & $ \lambda xy.y = KI $ & right projection \\\hline
% \end{tabular}

@@

\section{The category of retracts}\label{sec:retracts-category}

@@

\begin{remark}
  Hyland instead defines $ \R $ as
  \[ \R_0 = \{ A : L_1 \mid A \bullet A = A \} \quad \text{and} \quad \R(A, B) = \{ f : L_1 \mid B \bullet f \bullet A = f \}, \]
  writing $ s \bullet t $ instead of $ s \bullet (t) $ for $ s, t : L_1 $. Note that the following diagram commutes:
  \begin{center}
    \begin{tikzcd}
      L_1 \times L_1 \arrow[r, "\bullet"] \arrow[d, "{(s, t) \mapsto ((\lambda x_1, s), (\lambda x_1, t))}"'] & L_1 \arrow[d, "{s \mapsto (\lambda x_1, s)}"]\\
      L_0 \times L_0 \arrow[r, "\circ"] & L_0
    \end{tikzcd}
  \end{center}
  Now, since for $ A : \R $, $ A = \lambda x_1, \iota_{0, 1}(A) (\iota_{0, 1}(A) x_1) $, we have $ \lambda x_1, \iota_{0, 1}(A) x_1 = A $. This shows that $ s \mapsto \lambda x_1, s $ and $ t \mapsto \iota_{0, 1}(t) x_1 $ (both on objects and morphisms) constitute an equivalence between Hyland's category of retracts and Scott's category of retracts.
\end{remark}

@@

We might wonder whether $ \R $ also has an initial object $ O $. However, for any $ c: L_0 $, we would have a constant morphism
\[ I_c = \lambda x_1, \iota_{0, 1}(c) : \R(O, U), \]
so if $ L $ is nontrivial, $ \R $ has no initial object.

@@

$ \R $ also has exponential objects
\[ C^B = \lambda x_1, C \circ x_1 \circ B \]
with evaluation morphism $ \epsilon_{BC}: C^B \times B \to C $ given by
\[ \epsilon_{BC} = \lambda x_1, C(\pi_1 x_1 (B (\pi_2 x_1))), \]
which is universal because we can lift a morphism $ f: \R(A \times B, C) $ to a morphism $ \psi(f): \R(A, C^B) $ given by
\[ \psi(f) = \lambda x_1 x_2, f (x_1, x_2). \]
Note that for $ g: \R(A, C^B) $, the inverse $ \psi^{-1}(g) $ is given by
\[ \epsilon \circ \langle g \circ \pi_1, B \circ \pi_2 \rangle = \lambda x_1, g (\pi_1 x_1) (\pi_2 x_1): \R(A \times B, C). \]
Also note that
\[ \psi(\epsilon_{BC}) = \lambda x_1, C \circ x_1 \circ B = C^B = \id{C^B}. \]
Note that $ \varphi(C^B)_0 = \R(B, C) $, as we would expect.
